A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-26 imaging_demo/ms.tex
Restructure section 3, describe software modules
From 7ed89ad (master):
=====================

(*) Add new figure EPIC-modules.eps to the repository

In ms.tex,

(*) Add a new figure illustrating software architecture and modules
(*) Restructure subsections in section 3
(*) Add subsection on software architecture
(*) Move removal of auto-correlations to section 4 under Verification

The intermediate step of gridding with the antenna power response can be 
expressed as a convolution of a data vector generated by gridding the electric fields 
directly with the antenna illumination pattern.
\begin{align}
&\widetilde{\mathbf{B}}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v})=\nonumber\\
&\qquad\left\langle \left[\widetilde{\mathbf{W}}^\textrm{T}_a(\mathbf{r},\mathbf{a})\, \widetilde{\mathbf{N}}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}(\mathbf{a})\right]
\ast\left[\widetilde{\mathbf{W}}_a(\mathbf{r},\mathbf{a})\, \mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\, \mathbf{E}^\star(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

@@

@@

\subsection{Temporal Fourier transform}\label{sec:F-engine}

This module is common to the MOFF and FX imaging techniques. Time samples of
electric fields measured by the antenna and digitized by the A/D converter is
Fourier transformed to generate electric field spectra. This step can be
parallelized by antennas as shown in Fig.~\ref{fig:f-engine}. The output is
then fed to either MOFF and FX imaging pipelines.

@@

\subsection{Antenna-to-Grid Mapping}

A grid is generated on the coordinate system in which antenna locations are
specified with a grid spacing that is at most $\lambda_\textrm{min}/2$ even at the
highest frequency to ensure there is no aliasing even from regions of the sky far
away from the field of view. The number of locations on the grid is restricted to
be a power of 2.

@@

\subsection{Calibration}
Calibration of direct imaging correlators remains a challenge. Contrary
to the FX data flow, direct imagers mix the signals from all antennas before
averaging and writing to disk. It is therefore essential to apply gain solutions 
before the gridding step. Previous efforts have resorted to applying FX-generated
calibration solutions \citep{zhe14,fos14}, or integrating a dedicated FX 
correlator which periodically forms the full visibility matrix 
\citep{wij09,dev09}.

@@

The calibration module included in the EPIC repository allows for application of pre-
determined calibration solutions, or can solve for the complex gains using the 
EPICal algorithm.

\subsection{Gridding Convolution}

The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved by
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process using the sparse matrices module in Python's SciPy package. 
Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra,
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}).
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

\subsection{Spatial Fourier Transform}

Before the spatial Fourier transform, the gridded electric fields are padded with
zeros in order to match the grid size and angular size of each image pixel that
would have been obtained with software holography of output from an FX
correlator.

@@

\subsection{Time-averaging}

In MOFF imaging, the measured antenna electric fields and the corresponding
holographic electric field images are zero-mean stochastic quantities. Hence,
they cannot be time-averaged to reduce noise. The statistical quantity stable
with time in this case are the square of the holographic electric field images.
Thus, squared images have to be formed at every instant of time before averaging
as indicated in equation~\ref{eqn:dirty-image-MOFF}.

@@

@@

The squaring operation under MOFF imaging in the image plane introduces antenna 
auto-correlations around the zero spacing in the $uv$-plane which are absent in 
traditional visibility-based imaging. In order to facilitate a robust comparison, 
these auto-correlations are removed from both the $uv$ and image planes, which 
is otherwise not an essential part of the core algorithm. We describe below how 
they can be removed from the $uv$ plane in a straightforward manner.

@@

We inverse Fourier transform the squared 
images and beams back to the $uv$ plane and subtract the estimated 
auto-correlation kernel scaled to the peak value centered at the zero spacing 
pixel. The final averaged image is obtained by Fourier transforming the $uv$ 
plane data and weights with the auto-correlations subtracted to the image plane. 
These images are now comparable to those obtained from visibility-based imaging. 
This step of removing auto-correlations is required to be performed only once 
per integration timescale and does not add significant cost to the full operation.

@@

In the visibility-based imaging, the predominant computational cost is at
the X-engine requiring $N_\textrm{a}^2$ complex multiplications per channel 
per correlator writeout timescale.

In the following discussions, we will assume that the computational cost for
the MOFF imaging is determined by the spatial Fourier transform while  
that for visibility-based imaging comes from $N_\textrm{a}^2$ 
cross-correlations. However, if non-linearities such 
as non-coplanarity of baselines \citep{cor08} and wide-field phenomena like 
the {\it pitchfork} effect \citep{thy15a,thy15b} are to be corrected for, 
the antenna based illumination footprint can start becoming less compact
in the measurement plane and can result in a costlier gridding process.

@@

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of 
complex operations per frequency channel per integration timescale. Telescopes 
that fall to the left of this line indicate MOFF imaging is computationally
more efficient than visibility based imaging. All HERA layouts except possibly
HERA-19 are in a parameter space where MOFF imaging holds the advantage. The 
solid line showing future trajectory of HERA like systems will be clearly 
favoured by MOFF imaging. The gray shaded area is for a projected LWA 
expansion and is also predominantly in the region favouring MOFF imaging. 
It is bounded by the LWA1 and LWA-OV on the left and right 
respectively. The current (see Table~\ref{tab:antenna-layouts}) and
an expanded layout with a four-fold increase in number of elements over a
two-fold increase in $b_\textrm{max}$ provide the bounds at the bottom and top 
respectively. Current instruments such as MWA and LOFAR lie in parameter space
favouring visibility based imaging.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of
    number of complex multiplies and adds with MOFF and FX. The dashed line
    is the boundary at which the number of operations with MOFF and visibility 
    based imaging are equal. MOFF imaging is more efficient for telescopes 
    occupying the left of this line and vice versa. All the HERA layouts except
    possibly HERA-19 lie in the parameter space favoured by MOFF imaging. And so 
    are SKA1-LC and SKA1-LCD. The solid black curve shows the projected 
    trajectory of bigger close-packed hexagonal layouts similar to HERA. The 
    gray shaded area denotes the projected trajectory of the LWA bounded by LWA1 
    (left edge), LWA-OV (right edge), current layout (bottom) and a four-fold 
    increase in the number of elements in a two-fold increase in the core size
    (top). Current instruments such as MWA and LOFAR fall in a region favoured 
    by visibility based imaging.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}

>>>>

The intermediate step of gridding with the antenna power response can be 
expressed as a convolution of a data vector generated by gridding the electric fields 
directly with the antenna illumination pattern.
\begin{align}
&\widetilde{\mathbf{B}}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v})=\nonumber\\
&\qquad\left\langle \left[\widetilde{\mathbf{W}}^\textrm{T}_a(\mathbf{r},\mathbf{a})\, \widetilde{\mathbf{N}}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}(\mathbf{a})\right]
\ast\left[\widetilde{\mathbf{W}}_a(\mathbf{r},\mathbf{a})\, \mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}^\star(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

@@

\subsection{Data Flow}

EPIC can accept dual-polarization inputs and produce images of all four 
cross-polarizations. Currently two data input formats exist for reading in the 
electric field time samples measured by the antennas -- simulated electric 
fields based on a sky model using the simulator packaged with EPIC; and LWA 
data. Efforts to build interfaces for data from other telescopes are underway.

@@

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and FX imaging techniques. Time 
samples of electric fields measured by the antenna and digitized by the A/D 
converter is Fourier transformed to generate electric field spectra. This step 
can be parallelized by antennas as shown in Fig.~\ref{fig:f-engine}. The output 
is then fed to either MOFF and FX imaging pipelines.

@@

\par\medskip
\noindent {\bf Antenna-to-Grid Mapping}
\par\medskip
\noindent A grid is generated on the coordinate system in which antenna 
locations are specified with a grid spacing. The grid spacing can be controlled 
by the user. By default, it is set to be at $\lambda_\textrm{min}/2$ even at the 
highest frequency to ensure there is no aliasing even from regions of the sky 
far away from the field of view. The number of locations on the grid is 
restricted to be a power of 2.

@@

\par\medskip
\noindent {\bf Calibration}
\par\medskip
\noindent Calibration of direct imaging correlators remains a challenge. Contrary
to the FX data flow, direct imagers mix the signals from all antennas before
averaging and writing to disk. It is therefore essential to apply gain solutions 
before the gridding step. Previous efforts have resorted to applying FX-generated
calibration solutions \citep{zhe14,fos14}, or integrating a dedicated FX 
correlator which periodically forms the full visibility matrix 
\citep{wij09,dev09}.

@@

The calibration module included in the EPIC repository allows for application of 
pre-determined calibration solutions, or can solve for the complex gains using 
the EPICal algorithm.

\par\medskip
\noindent {\bf Gridding Convolution}
\par\medskip
\noindent The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the 
individual antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved by
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process using the sparse matrices module in Python's SciPy package. 
Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra,
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}).
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

\par\medskip
\noindent {\bf Spatial Fourier Transform}
\par\medskip
\noindent Before the spatial Fourier transform, the gridded electric fields are 
padded with zeros in order to match the grid size and angular size of each image 
pixel that would have been obtained with software holography of output from an FX
correlator.

@@

\par\medskip
\noindent {\bf Time-averaging}
\par\medskip
\noindent In MOFF imaging, the measured antenna electric fields and the 
corresponding holographic electric field images are zero-mean stochastic 
quantities. Hence, they cannot be time-averaged to reduce noise. The statistical 
quantity stable with time in this case are the square of the holographic 
electric field images. Thus, squared images have to be formed at every instant 
of time before averaging as indicated in equation~\ref{eqn:dirty-image-MOFF}.

@@

\subsection{Software Architecture}\label{sec:software-modules}

EPIC is built using object oriented programming in Python and is built on
carefully crafted modules which closely represent real-life entities in radio 
interferometer arrays and observations. The essential modules along with their
key attributes and methods are illustrated in Fig.~\ref{fig:software-modules}.
These modules are described below.

\begin{figure*}
  \includegraphics[width=\linewidth]{EPIC-modules.eps}
  \caption{Software architecture of EPIC with core modules, their essential
    attributes and functions. The Antenna module forms the fundamental building 
    block. It consists of electric field time-series and spectra and the 
    F-engine that performs a temporal FFT to obtain electric field spectra from 
    the time-series. The Interferometer module is made of a pair of Antenna 
    modules. Its main function is the X-engine (FX or XF) to produce visibility 
    spectra. The Antenna Array module is made of all individual Antenna modules
    as its components and contains collective properties about the antenna 
    subsystems. Its core function is the creation of antenna-to-grid mapping and 
    that of gridded aperture weights and electric fields. The Interferometer 
    Array module is very similar in principle to the Antenna Array module 
    except it operates on cross-correlations and produces gridded visibilities.
    The Image module takes gridded electric fields or visibilities and performs
    a two-dimensional spatial FFT (and squares the intermediate image in case 
    of the former) to produce output images.}
  \label{fig:software-modules}
\end{figure*}

\subsubsection{Antenna module}

The Antenna module is a fundamental building block upon which all the other 
modules are built. There is one Antenna module per antenna each having 
attributes -- the propagated electric field time-series, $E(t)$, and spectrum 
$E(f)$ for both polarizations. The most important function inside this module 
is the F-engine that Fourier transforms time-series electric field data into 
spectrum.

The other function (not shown in the figure) is to update the data as new data 
keeps streaming in. This can also be parallelized. Another important attribute
(not shown in figure) are antenna flags for each polarization appropriate for 
the data stream being held by the Antenna module.

\subsubsection{Interferometer module}

The Interferometer module holds the attributes and functions pertaining to
a pair of antennas and represents the cross-correlation information obtained
from the pair. Its important attributes are the two Antenna modules. It also
contains four cross-polarized visibility time-series and spectra.

The most important function of the Interferometer module is the X-engine. 
This is essentially a software analog of hardware correlators of real 
telescope systems. The X-engine can be toggled between two states of operation,
namely, the FX and XF modes. The FX mode obtains the electric field spectra, 
$E(f)$ from the individual Antenna modules inside the Interferometer module and
multiplies the two to obtain visibility spectra, $V(f)$. On the other hand, the
XF mode cross-correlates the electric field time-series from its Antenna modules
to obtain the visibilities as a function of lags, $V_t(t)$, which is then Fourier 
transformed to obtain $V(f)$. Both modules can operate on dual-polarizations to
obtain all four cross-polarizations.

The other attributes (not shown in figure) are the flags applicable for each
cross-polarization for the current data stream. Similar to the Antenna module,
it has an update function that can update the visibilities $V_t(t)$ or $V(f)$
directly rather than through the electric fields of its component antennas. 
This functionality is to allow EPIC to operate while attached to the backend 
of traditional correlator systems. Currently, this feature is not utilized 
for purposes of this paper.

The Interferometer module forms the fundamental unit for the InterferometerArray
module (to be discussed below) and in general for visibility based correlator 
and imaging systems.

\subsubsection{Antenna Array module}

The AntennaArray module consists of all the Antenna modules as its attributes
and represents the collective properties of its component antennas. By virtue of 
holding each antenna data independently in their respective modules the F-engine 
for the entire array can be distributed to the F-engines of the component 
antenna modules thus achieving a highly parallelized F-engine while emulating 
real telescope systems.

The most important attributes held by this module are the antenna aperture 
illumination weights and electric fields projected on the grid using the 
gridding convolution method described above and implemented by the gridding 
function (denoted by grid()) in this module. Significant parts of the 
antenna-to-grid mapping and gridding convolution are parallelizable 
across antennas and frequencies.

Individual antenna flags are carried over as additional weights to be applied
to the gridded aperture illumination and electric fields. A series of data 
streams can be stacked up to take advantage of the array optimization available 
in Python. This module is also equipped to manage dual-polarization.

\subsubsection{Interferometer Array module}

Similar to the AntennaArray module, the InterferometerArray module consists of
individual Interferometer modules. It can parallelize the correlator operations
by distributing the X-operation over the X-engines of its component 
Interferometer modules. The interferometer-to-grid mapping and gridding 
convolution are very similar in nature to that of the Antenna Array module. 
Flag-based grid weights, stacking and ability to handle all four 
cross-polarizations are built into this module.

\subsubsection{Image module}

The Image module is built as a general purpose module that can switch between 
operating on gridded electric fields or visibilities. At its heart, it consists
of a two-dimensional FFT where the padding can be specified by the user to 
control the resolution in the output images. In case of electric field based
imaging, there is an additional step of squaring the holographic electric field
images. This is packaged into the function annotated as imagr().

Besides its core functions of spatial Fourier transform and squaring, it can 
stack, accumulate and average images, and optionally remove the antenna
auto-correlations centered around the zero-spacing pixel in the $uv$ plane. 
It also handles all four cross-polarization products. Currently, it can 
support writing data out in standard FITS format.

@@

Before diving into the details of simulations and comparison with visibility 
based imaging, we describe the elimination of a minor mathematical difference 
between the two techniques. The squaring operation under MOFF imaging in the 
image plane introduces antenna auto-correlations around the zero spacing in 
the $uv$-plane which are absent in traditional visibility-based imaging. In 
order to facilitate a robust comparison, these auto-correlations are removed 
from both the $uv$ and image planes, which is otherwise not an essential part 
of the core algorithm. We describe below how they can be removed from the $uv$ 
plane in a straightforward manner.

@@

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centered at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility-based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration timescale and does not add significant cost to the full operation.

@@

In the visibility-based imaging, the predominant computational cost is at
the X-engine requiring $N_\textrm{a}(N_\textrm{a}-1)/2$ complex multiplications 
per channel at every correlator writeout timescale.

In the following discussions, we will assume that the computational cost for
the MOFF imaging is determined by the spatial Fourier transform while  
that for visibility-based imaging comes from $N_\textrm{a}(N_\textrm{a}-1)/2$ 
cross-correlations. However, if non-linearities such as non-coplanarity of 
baselines \citep{cor08} and wide-field phenomena like the {\it pitchfork} 
effect \citep{thy15a,thy15b} are to be corrected for, the antenna based 
illumination footprint can start becoming less compact in the measurement 
plane and can result in a costlier gridding process.

@@

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of 
complex operations per frequency channel per integration timescale. Telescopes 
that fall to the left of this line indicate MOFF imaging is computationally
more efficient than visibility based imaging. All HERA layouts except possibly
HERA-19 are in a parameter space where MOFF imaging holds the advantage. The 
solid line showing future trajectory of HERA like systems will be clearly 
favoured by MOFF imaging. The gray shaded area is for a projected LWA 
expansion and is also predominantly in the region favouring MOFF imaging. 
It is bounded by the LWA1 and LWA-OV on the left and right 
respectively. The current (see Table~\ref{tab:antenna-layouts}) and
an expanded layout with a four-fold increase in number of elements over a
50\% increase in $b_\textrm{max}$ provide the bounds at the bottom and top 
respectively. Current instruments such as MWA and LOFAR lie in parameter space
favouring visibility based imaging.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_computations_fov_gridding_annotated.eps}
  \caption{Current and planned instruments in parameter space of
    number of complex multiplies and adds with MOFF and FX. The dashed line
    is the boundary at which the number of operations with MOFF and visibility 
    based imaging are equal. MOFF imaging is more efficient for telescopes 
    occupying the left of this line and vice versa. CHIME, HIRAX and all the 
    HERA layouts except HERA-19 and HERA-37 lie in the parameter space favoured 
    by MOFF imaging. And so are SKA1-LC and SKA1-LCD. The solid black curve 
    shows the projected trajectory of bigger close-packed hexagonal layouts 
    similar to HERA. The gray shaded area denotes the projected trajectory of 
    the LWA bounded by LWA1 (left edge), LWA-OV (right edge), current layout 
    (bottom) and a four-fold increase in the number of elements within a 50\% 
    increase in the core size (top). Current instruments such as MWA and LOFAR 
    fall in a region favoured by visibility based imaging.}
  \label{fig:parameter-space-computations-instruments}
\end{figure}
